---
title: "Week 5 Temporal analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F,message = F)
```

## Getting to UFApps

Here are the instructions to get to UFApps - they are the same as last week. 
\

Navigate to: https://info.apps.ufl.edu

Click on the orange "Login to UFApps" button in the upper right. This link may also take you to the login page: https://apps.ufl.edu/ 

Log in using your GatorLink.

Click on "Apps" in the top menu bar. Scroll down to RStudio and click on it to launch it. It may take a minute. Sometimes the connection will not go through, but if you launch it again it works. If you consistently can't get through, reach out to Renata and she will try and figure out another solution. 

You may see a prompt to "set up secure access to your files". There's no need to do this for this course, so go ahead and click Not Now. 

## Setup

Select "File", "New File", and "RScript" in the top left menu bar. This will give you a place to copy and paste code.

To run code, either highlight it and click the "Run" button in the top right of the top left pane, or highlight it and press Ctrl + Enter. 

Copy and paste this code to set up. You may get a notification about copy-pasting stuff into the virtual session. Click OK and continue.

```{r setup dirs, eval = F}


if(!dir.exists("Documents/EVS3000L")) {
dir.create("Documents/EVS3000L")
}
if(!dir.exists("Documents/EVS3000L/week5")) {
dir.create("Documents/EVS3000L/week5")
}
if(!dir.exists("Documents/EVS3000L/week5/data")) {
dir.create("Documents/EVS3000L/week5/data")
}
setwd("Documents/EVS3000L/week5/")

library(ggplot2)

```

## Download data

We will download data on a community of desert rodents located in southeastern Arizona. This community has been monitored on a monthly basis for over 40 years. 


```{r get portal data with portalr, include = F}

library(dplyr)


raw_portal_data <- portalr::abundance(time = "date")

abundance_ts <- data.frame(
  censusdate = raw_portal_data$censusdate,
  total_abund = rowSums(raw_portal_data[, 2:22])
)

richness_ts <- data.frame(
  censusdate = raw_portal_data$censusdate,
  richness = rowSums(raw_portal_data[,2:22] > 0)
)

spectab_ts <- raw_portal_data %>%
  select(censusdate, DS)


ts <- left_join(abundance_ts, richness_ts) %>%
  left_join(spectab_ts)


ts <- ts %>%
  mutate(year = as.character(censusdate)) %>%
  mutate(year = substr(year, 0, 4)) %>%
  mutate(year = as.numeric(year)) %>%
  filter(year > 1980) %>%
  group_by(year) %>%
  summarize(total_abundance = sum(total_abund),
            richness = sum(richness),
            spectab = sum(DS))

write.csv(ts, "data/rodent_data.csv", row.names = F)

rodent_data <- ts

```

```{r get portal data from repo, eval = F}

download.file("https://raw.githubusercontent.com/diazrenata/EVS3000L-fall2020/master/week5_temporal/data/rodent_data.csv", destfile = "data/rodent_data.csv")

rodent_data <- read.csv("data/rodent_data.csv", stringsAsFactors = F)

```

## Plot abundance and species richness data

We will look at how two aspects of this community have changed over the past 40 years: the total number of species present, or species richness, and the total number of *individuals* present, or total abundance.

```{r plot abundance and richness}
library(ggplot2)

richness_plot <- ggplot(rodent_data, aes(x = year, y = richness)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  ggtitle("Species richness over time")

richness_plot


abundance_plot <- ggplot(rodent_data, aes(x = year, y = total_abundance)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  ggtitle("Total number of individuals over time")

abundance_plot


spectab_plot <- ggplot(rodent_data, aes(x = year, y = spectab)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  ggtitle("Dipodomys spectabilis over time")

spectab_plot

```

**What dynamics do you notice in these three time series? Do they look like they are increasing, decreasing, or not changing?** Include your observations from these plots in your text submission. 

## Statistical analysis

We will use a basic statistical test to see whether we can detect evidence that these timeseries are changing. Specifically, we will fit a **linear model**, which will tell us whether the line-of-best-fit for the timeseries has a slope that is statistically distinguishable from 0. A slope of 0 reflects no systematic change over time. 

[fit lm]

[plot fit]

[print results]

**What do the linear models tell us about change in species richness and abundance over the time series?**


## Reflections

Finish your text submission by answering these questions:

- Whenever we use a statistical model, we need to think critically about whether it is the appropriate tool for the question we are interested in. In this case, do you think the linear model is accurately describing what you see in the data? Is there something else you might want to try and capture with a more complicated or nuanced model?
- Last week we saw a little bit of what kinds of spatial and environmental data we can access from other sources. Can you brainstorm questions you might be able to answer by combining spatial environmental data with timeseries data like what we have here?


## Results to submit

Save your plots and results with this code:

Submit your plots and results to eLearning. [elearning upload instructions]