---
title: "Week 5 Temporal analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F,message = F)
```

## Getting to UFApps

Here are the instructions to get to UFApps - they are the same as last week. 
\

Navigate to: https://info.apps.ufl.edu

Click on the orange "Login to UFApps" button in the upper right. This link may also take you to the login page: https://apps.ufl.edu/ 

Log in using your GatorLink.

Click on "Apps" in the top menu bar. Scroll down to RStudio and click on it to launch it. It may take a minute. Sometimes the connection will not go through, but if you launch it again it works. If you consistently can't get through, reach out to Renata and she will try and figure out another solution. 

You may see a prompt to "set up secure access to your files". There's no need to do this for this course, so go ahead and click Not Now. 

## Setup

Select "File", "New File", and "RScript" in the top left menu bar. This will give you a place to copy and paste code.

To run code, either highlight it and click the "Run" button in the top right of the top left pane, or highlight it and press Ctrl + Enter. 

Copy and paste this code to set up. You may get a notification about copy-pasting stuff into the virtual session. Click OK and continue.

```{r local setup, include = F}

library(ggplot2)
library(dplyr)

```

```{r setup dirs, eval = F}


if(!dir.exists("Documents/EVS3000L")) {
dir.create("Documents/EVS3000L")
}
if(!dir.exists("Documents/EVS3000L/week5")) {
dir.create("Documents/EVS3000L/week5")
}
if(!dir.exists("Documents/EVS3000L/week5/data")) {
dir.create("Documents/EVS3000L/week5/data")
}
setwd("Documents/EVS3000L/week5/")

library(ggplot2)
library(dplyr)

```

## Download data

We will download data on a community of desert rodents located in southeastern Arizona. This community has been monitored on a monthly basis for over 40 years. 


```{r get portal data with portalr, include = F, eval = F}

library(dplyr)


raw_portal_data <- portalr::abundance(time = "date")

abundance_ts <- data.frame(
  censusdate = raw_portal_data$censusdate,
  total_abund = rowSums(raw_portal_data[, 2:22])
)

richness_ts <- data.frame(
  censusdate = raw_portal_data$censusdate,
  richness = rowSums(raw_portal_data[,2:22] > 0)
)

spectab_ts <- raw_portal_data %>%
  select(censusdate, DS, PP, DM, PB)


ts <- left_join(abundance_ts, richness_ts) %>%
  left_join(spectab_ts)


ts <- ts %>%
  mutate(year = as.character(censusdate)) %>%
  mutate(year = substr(year, 0, 4)) %>%
  mutate(year = as.numeric(year)) %>%
  filter(year > 1980) %>%
  group_by(year) %>%
  summarize(total_abundance = sum(total_abund),
            richness = sum(richness),
            bannertail = sum(DS),
            merriami = sum(DM),
            pocketmouse = sum(PP),
            baileys = sum(PB))

write.csv(ts, "data/rodent_data.csv", row.names = F)

rodent_data <- ts

```

```{r get portal data from repo, eval = T}

download.file("https://raw.githubusercontent.com/diazrenata/EVS3000L-fall2020/master/week5_temporal/data/rodent_data.csv", destfile = "data/rodent_data.csv")

rodent_data <- read.csv("data/rodent_data.csv", stringsAsFactors = F)

```

## Abundance and species richness data

We will look at how two aspects of this community have changed over the past 40 years: the total number of species present, or species richness, and the total number of *individuals* present, or total abundance.

**The questions in this section are for you to mull over as you work through. You do not need to answer them in your submission. When we get to analyzing specific species, please do answer those questions in your submission.**

```{r plot abundance and richness, eval = T}

richness_plot <- ggplot(rodent_data, aes(x = year, y = richness)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  ggtitle("Species richness over time")

richness_plot


abundance_plot <- ggplot(rodent_data, aes(x = year, y = total_abundance)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  ggtitle("Total number of individuals over time")

abundance_plot
```

**What dynamics do you notice in these two time series?**

* Do they look like they are increasing, decreasing, or not changing? 
* Are they tightly bounded around a consistent value, or variable? Is this consistent over time?
* Does species richness (diversity) seem to match total abundance?


## Statistical analysis on abundance and species richness

While we can get a general intuition about what our data are saying by simply looking at the plots, it's helpful to use a statisical analysis to test for dynamics that might not be obvious to the human eye, or to test whether the patterns that we *think* we see are in fact statistically distinguishable from randomness! (Humans are very good at coming up with patterns when presented with data, even if the data is in fact random noise).

Here, we will use ordinary **linear models** to estimate the slope, or rate of change, for species richness and total abundance over time. We will also test whether the slope is **statistically significant**, or if our model is unable to detect a slope that is statistically different from 0. Linear models are a common entry point for statistical analysis, especially of temporal data (although there are, not surprisingly, many variations that can get considerably more nuanced and more complex!)


### Species richness

Let's fit a linear model to estimate how species richness has changed over time:

```{r richness lm}

richness_lm <- lm(richness ~ year, data = rodent_data)

summary(richness_lm)

```


R has told us lots of things about our linear model. We'll focus on the **statisical significance**, or p-value, of the model and the **estimated slope**. 

The **p-value** for this model is .212. Conventionally, if the p-value for a model is greater than .05, we **do not consider the model statistically significant**. In this case, this means we are not confident that our data have a slope that is different from 0. 

We will still note the estimated slope. It is the "Estimate" for "year" in the summary output above. For this model, is -.56. However, because the p-value is much larger than .05, we can't be sure that the slope isn't really 0. 

Another way to think of this is that the confidence intervals for the estimate of slope overlap 0:

```{r confints}

confint(richness_lm)

```


So, we cannot say with confidence that species richness is changing systematically, either up or down. 

**Does this match your intuition from looking at the plot?**


### Total abundance

Let's repeat this analysis, using total abundance. 

```{r abund lm}

abund_lm <-  lm(total_abundance ~ year, data = rodent_data)

summary(abund_lm)

confint(abund_lm)

```

In contrast to the species richness model:

* The p-value for total abundance is .033, which is less than .05!
* The confidence intervals for the slope do not overlap 0
* The estimate for the slope is 24.7

We therefore have greater confidence that our estimated slope reflects a real signal in the data, and is not random noise. In this case, the slope is positive, meaning that total abundance has increased over the course of the time series.

**How does this correspond to your intuition from the plot?**


## Working with a particular species

Now we will look at some of the particular species we have been monitoring at this site. We will work through an analysis for one species, and then you will select another species to analyze on your own.

**Please answer the questions in this section in a Word or text document and submit them along with your results.** The questions you should answer are also collected here: 


### Banner-tailed kangaroo rats

*Dipodomys spectabilis*, or the banner-tailed kangaroo rat, is one of our most readily identified study species. It's the largest kangaroo rat we find at our site - it's about the size of a sweet potato, not counting the tail. It's called a banner-tailed kangaroo rat because its tail has a magnificient white tuft that looks like a banner behind it. It's an important player in our system, because it eats the seeds of many of the desert plants, and it is behaviorally dominant to other, smaller, rodents. Wikipedia has some good images and natural history: https://en.wikipedia.org/wiki/Banner-tailed_kangaroo_rat


Here we will plot how the abundance of *D. spectabilis* has changed over time:

```{r plot spectabs, eval = F}

bannertail_plot <- ggplot(rodent_data, aes(x = year, y = bannertail)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  ggtitle("Bannertail kangaroo rats over time")

bannertail_plot

```


**What dynamics stand out to you**? Changes in abundance over time? Steadiness, or variability?


Let's fit a linear model to get a more quantitative description of how bannertail abundance has changed over time:

```{r spectab lm, eval = F}

bannertail_lm <- lm(bannertail ~ year, data = rodent_data)

summary(bannertail_lm)

confint(bannertail_lm)

```


From these results, answer these questions as part of your submission: 

* What is the slope estimate for the bannertails?
* Is this model statistically significant? (p value < .05, or the confidence intervals for the slope estimate **not** overlapping 0)
* Does this fit with your intuition from the plot?
* If you got these results for a species you were monitoring, what would you think? What questions or studies would it prompt you to look into further? 


### Choose your own species

Now you will do the same analysis as we've done for species richness, abundance, and bannertail kangaroo rats with a species of your choosing from our site. Here are your options:

* Merriam's kangaroo rat, *Dipodomys merriami*. Merriam's are smaller than bannertails - about the size (and shape) of two ping-pong balls stuck together and covered in fur. They also lack the banner-tail, although they do have tail tufts. Wikipedia: https://en.wikipedia.org/wiki/Merriam%27s_kangaroo_rat
* Desert pocket mouse, *Chaetodipus penicillatus*. While the kangaroo rats hop around on their back two legs, like little kangaroos, the pocket mice are quadripedal - more like your sterotypical mouse. Like kangaroo rats, they have cheek pouches that they use to carry seeds back to their underground burrows. *C. penicillatus* is the smallest of the species included here. They vary in size from a little bigger than a tube of Chapstick to closer to two tubes of Chapstick. Wikipedia: https://en.wikipedia.org/wiki/Desert_pocket_mouse 
* Bailey's pocket mouse, *Chaetodipus baileyi*. Bailey's appear very similar to desert pocket mice, but about 2/3 larger. Wikipedia: https://en.wikipedia.org/wiki/Bailey%27s_pocket_mouse

To run the analysis using your chosen species, replace `bannertail` in the following lines of code with `merriami`, `pocketmouse`, or `baileys` depending on which species you'd like to use.

First, plot how your species has changed over time:

```{r plot your species, eval = F}

my_species_plot <- ggplot(rodent_data, aes(x = year, y = bannertail)) + # Replace bannertail with your species
  geom_point() +
  geom_line() +
  theme_bw() +
  ggtitle("My species' abundance over time")

my_species_plot

```


What do the dynamics for your species *look like* to you? Would you describe them as increasing, decreasing, random, or something else?

Now fit a linear model to quantitively describe the dynamics:

```{r lm your species, eval = F}

my_species_lm <- lm(bannertail ~ year, data = rodent_data) # Replace bannertail with your species

summary(my_species_lm)

confint(my_species_lm)

```


* Is this model statistically significant?
* If it is, what is the slope estimate? Is this species increasing or decreasing over time?
* Does this species' dynamics match or differ from the bannertail kangaroo rats? 
* If you were monitoring this species, what questions would these results inspire?

## Reflection questions

* How do the dynamics of individual species correspond to the overall dynamics of species richness and total abundance? 
* In your opinion, do the linear models tend to match your sense of the dynamics you see from the plots? Are there patterns you see that the models don't seem to capture? 
* What questions do these results inspire? 
* What additional data or information could you imagine gathering that would help you better understand these results?

## Results to submit

Save your plots and results with this code:

```{r save results, eval =F}


jpeg('species_dynamics.jpg')
plot(my_species_plot)
dev.off()

saveRDS(my_species_lm, file = "species_lm.Rds")

```


Submit `species_dynamics.jpg` and `species_lm.Rds` to eLearning via file upload, along with a Word or text document with your answers to the questions. To help you keep track of the questions, the ones you should answer in your submission are listed here: 


