---
title: "Week 4 Spatial analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F,message = F, eval = F)
```

## Getting to UFApps

We will use a version of R that is hosted in the cloud for UF students through UFApps.

Navigate to: https://info.apps.ufl.edu

Click on the orange "Login to UFApps" button in the upper right. This link may also take you to the login page: https://apps.ufl.edu/ 

Log in using your GatorLink.

Click on "Apps" in the top menu bar. Scroll down to RStudio and click on it to launch it. It may take a minute. Sometimes the connection will not go through, but if you launch it again it works. If you consistently can't get through, reach out to Renata and she will try and figure out another solution. 

You may see a prompt to "set up secure access to your files". There's no need to do this for this course, so go ahead and click Not Now. 

## Setup

Select "File", "New File", and "RScript" in the top left menu bar. This will give you a place to copy and paste code.

To run code, either highlight it and click the "Run" button in the top right of the top left pane, or highlight it and press Ctrl + Enter. 

Copy and paste this code to set up. You may get a notification about copy-pasting into the virtual session. Click OK and continue.


This code will set up your file directories so you will be able to download data files and save your results later:


```{r install packages, include =TRUE, eval= F}


if(!dir.exists("Documents/EVS3000L")) {
dir.create("Documents/EVS3000L")
}
if(!dir.exists("Documents/EVS3000L/week4")) {
dir.create("Documents/EVS3000L/week4")
}
if(!dir.exists("Documents/EVS3000L/week4/prism")) {
dir.create("Documents/EVS3000L/week4/prism")
}
setwd("Documents/EVS3000L/week4/prism")

```


This code will download the activate the tools we need to work with spatial data in R. (R is an especially powerful programming environment because R users all over the world volunteer to build __R packages__, which are custom-built sets of functions to accomplish all kinds of analyses and other tasks. R packages are free and open-source, which means anyone can download them and use them for their work. Here we are downloading and activating several R packages that are commonly used for spatial analysis.) 

```{r libraries}
install.packages(c("prism", "tigris", "raster", "rgdal"))


library(dplyr)
library(ggplot2)
library(prism)
library(tigris)
library(rgdal)
```

You may get "warning messages". That's okay.


Here we'll tell R where to put the climate data we're about to download:


```{r prism path local, include = F}
options(prism.path = here::here("week4_spatial", "prism"))
```

```{r prism path ufapps, eval = F}
options(prism.path = getwd())
```

## Download and load temperature and precipitation files

We'll download spatial data files of *annual precipitation* and *mean annual temperature* for the contiguous United States in 1985 and 2014 from the Prism Climate Project. This code will download the data files to your files:


```{r download prism files, eval =F}
get_prism_annual(type="ppt", years = c(1985, 2014), keepZip=FALSE)
get_prism_annual(type="tmean", years = c(1985, 2014), keepZip=FALSE)
```


And this code will load the data files you just downloaded into your R session:


```{r load prism}
prism_files <- ls_prism_data()
prism_dat <- prism_stack(prism_files)
names(prism_dat) <- c("precip_1985", "precip_2014", "temp_1985", "temp_2014")

```

## Plot temperature and precipitation data

It's good practice to look at data whenever you first load it. We'll plot these data as heatmaps of precipitation and temperature:

```{r plot prism}

plot(prism_dat)

```

## Compute variable change

Now let's explore some ways to analyze this data. We'll calculate how much these climate variables have changed from 1985 until 2014. We'll subtract the 1985 values from the 2014 values, and plot the heatmaps of *change*:

```{r compute change}

precip_change = prism_dat$precip_2014 - prism_dat$precip_1985
temp_change = prism_dat$temp_2014 - prism_dat$temp_1985

plot(precip_change)
plot(temp_change)

```

## State boundaries


Say we are interested in how climate has changed in a particular state over this period. We can filter the PRISM climate data to focus on a particular geographic area. To filter to a specific state, we need the state boundaries. 

The US Census department publishes quite a lot of useful data products, including maps of state boundaries. This code will load the census map of state boundaries in to R, and filter it to remove the non-contiguous states and territories (which are not included in the PRISM climate data).

```{r get state boundaries}
states <- states(cb = T)

states <- states[, "NAME"]

states <- filter(states,
                   NAME != "United States Virgin Islands",
              NAME != "Commonwealth of the Northern Mariana Islands",
              NAME != "Hawaii",
              NAME != "Guam",
              NAME != "American Samoa",
              NAME != "Alaska",
              NAME != "Puerto Rico")

```


Let's plot what we have just loaded. 


```{r plot contiguous}
plot(states)

```

We have the boundaries for all the states and territories. **We'll work through the next few chunks of code focusing on Florida. Then, re-do this analysis yourself using a different state.**


## Extracting Florida

This will extract just the boundaries for Florida:

```{r choose Florida}

florida_boundaries <- filter(states, NAME == "Florida")

plot(florida_boundaries)
```


## Plot precipitation and temperature change for Florida

We use the `crop` and `mask` functions to filter our maps of temperature and precipitation change for the whole US, to just Florida:

```{r plot Florida change}
library(raster)

florida_temp_change <- crop(mask(temp_change, florida_boundaries[1,]), florida_boundaries[1,])

florida_precip_change <- crop(mask(precip_change, florida_boundaries[1,]), florida_boundaries[1,])

plot(florida_temp_change)

plot(florida_precip_change)

```

## Compute mean variable change for Florida

We can extract the actual data values and compute summary statistics on them. For example, we can compute the mean temperature change and mean precipitation change across the whole state:

```{r mean Florida change}

florida_mean_temp_change =extract(florida_temp_change, florida_boundaries[1,], fun = mean, na.rm = T)

florida_mean_precip_change =extract(florida_precip_change, florida_boundaries[1,], fun = mean, na.rm = T)

print(paste0("Mean temperature change: ", florida_mean_temp_change))

print(paste0("Mean precipitation change: ", florida_mean_precip_change))


```

## Run with a diffferent state

Now you'll repeat the filtering and summary statistic computation using a state other than Florida. To do this, run the following code and replace `NAME == "Florida"` below with `NAME == "(the name of the state you choose)"`. 

```{r choose a state}

my_state <- filter(states, NAME == "Florida")

plot(my_state)
```

**You must spell the name of your state, and capitalize it, exactly as it is spelled in the data file R has!** This will give you a list of exactly how everything is spelled in the data file. 

```{r state options}

unique(states$NAME)

```


## Plot precipitation and temperature change for your  state

```{r plot state change}

my_state_temp_change <- crop(mask(temp_change, my_state[1,]), my_state[1,])

my_state_precip_change <- crop(mask(precip_change, my_state[1,]), my_state[1,])

plot(my_state_temp_change)

plot(my_state_precip_change)

```

## Compute mean variable change for your state

```{r mean change}

mean_temp_change =extract(my_state_temp_change, my_state[1,], fun = mean, na.rm = T)

mean_precip_change =extract(my_state_precip_change, my_state[1,], fun = mean, na.rm = T)

print(paste0("Mean temperature change: ", mean_temp_change))

print(paste0("Mean precipitation change: ", mean_precip_change))


```


## Results to submit

Running this code will save copies of your plots and results so you can upload them to eLearning. 

```{r results}

jpeg('temp_change.jpg')
plot(my_state_temp_change)
dev.off()


jpeg('precip_change.jpg')
plot(my_state_precip_change)
dev.off()

results <- data.frame(
  state = my_state$NAME[1],
  mean_temp_change = mean_temp_change,
  mean_precip_change = mean_precip_change
)

write.csv(results, "results.csv", row.names = F)

```

After you run this section, go ahead and exit RStudio.

## Uploading your results

Go back to the Apps page in UFApps. 

Scroll to eLearning and launch it. Log in to Canvas.

Navigate to the Spatial Analysis assignment and go to Submit and File Upload. 

In the file selection navigation window, go to Documents > EVS3000L > week4 > prism. Select "precip_change.jpg", "temp_change.jpg", and "results.csv". Upload them and submit. 

## Reflections

Please also upload a Word or text document with your answers to these questions about the exercise:

* How do the map, direction, or magnitude of change in your chosen state compare or contrast with your intuition or previous expectations?
* Are there other variables you think might give you a more informative or complete picture?
* More broadly, can you think of ways you could combine spatial information, like what we've used here, with the local biodiversity sampling you have done so far to learn more about the state of ecosystems locally or more generally?
